---
title: "Quick NAFE Stats"
author: "Brittny Vongdara"
date: 2023-04-19
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = file.path(dirname(inputFile), 'index.html')) }) 
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, fig.width = 10)

# Read in libraries ----
library(rvest)
library(tidyr)
library(stringr)
library(dplyr)
library(ggplot2)
library(plotly)
library(scales)
library(MetBrewer)

# Functions ----
remove_spaces <- function(spacey_cols) {
  gsub(pattern = "\\s+",
       replacement = " ",
       x = spacey_cols)
}

remove_commas <- function(decomma_cols) {
  as.numeric(gsub(
    pattern = ",",
    replacement = "",
    x = decomma_cols
  ))
}

# Gather and clean data ----
# Define root directory for TBS website with salary data
url <-
  "https://www.canada.ca/en/treasury-board-secretariat/services/innovation/human-resources-statistics/diversity-inclusion-statistics/"

temp_df <- 
  html_nodes(read_html(paste0(
    url,
    "distribution-public-service-canada-employees-designated-sub-group-department-members-visible-minorities"
  )), "table") %>%
  html_table(fill = TRUE)

row_no <- -c(1, 70:71) # there are 69 departments from 2021-2019 and 68 departments prior to 2019
col_no <- c(1, 2, 9, 11, 13, 15, 17, 21, 25)

df <-
  bind_rows({
    temp_df[[1]][row_no, col_no] %>%
      mutate(year = 2021)
  }, {
    temp_df[[2]][row_no, col_no] %>%
      mutate(year = 2020)
  },
  {
    temp_df[[3]][row_no, col_no] %>%
      mutate(year =  2019)
  },
  {
    temp_df[[4]][row_no, col_no] %>%
      mutate(year = 2018)
  },
  {
    temp_df[[5]][row_no, col_no] %>%
      mutate(year = 2017)
  }) %>% 
  filter(!`Department / Agency` %in% c("Total", ""),
         !str_detect(`Department / Agency`, "Table 4")) %>% 
  mutate(
    across(!`Department / Agency`,
           ~ replace(., . ==  "Table 1 Footnote *" ,
                     NA)), # replace footnote with NA
    across(!`Department / Agency`, remove_commas),
    `Department / Agency` = gsub(
      pattern = "\\s+",
      replacement = " ",
      x = `Department / Agency`
    ),
    `Department / Agency` = gsub(
      pattern = "Table [0-9] Footnote [0-9]",
      replacement = "",
      x = `Department / Agency`
    ),
    `Department / Agency` = trimws(`Department / Agency`),
    asian_pop = rowSums(across(
      c(
        Chinese,
        Japanese,
        Korean,
        Filipino,
        `South Asian/ East Indian`,
        `Southeast Asian`
      )
    ), na.rm = TRUE),
    sub_asian_all = `All employees` - asian_pop,
    pct_vm = round(asian_pop/`Members of visible minorities` * 100, 2),
    pct_all = round(asian_pop/`All employees` * 100, 2),
    pct_non_asian = round(sub_asian_all/`All employees` * 100, 2),
  ) %>%
  pivot_longer(cols = c(
    `All employees`,
    `Members of visible minorities`,
    asian_pop,
    sub_asian_all,
    Chinese,
    Japanese,
    Korean,
    Filipino,
    `South Asian/ East Indian`,
    `Southeast Asian`
    # pct_all,
    # pct_non_asian
    # pct_vm
  ),
  names_to = "group",
  values_to = "population")

## Asian population ----
asian_pop <- 
  df %>% 
  filter(group %in% c("sub_asian_all", "asian_pop")) %>% 
  group_by(year, group) %>% 
  summarise(total = sum(population)) %>% 
  ungroup() %>% 
  mutate(group = ifelse(group == "asian_pop", "Asian population", "Non-Asian population"))

yoy <- 
  asian_pop %>% 
  group_by(group) %>% 
  mutate(yoy = round((total - lag(total)) / lag(total) * 100, 2)) %>% 
  ungroup()

## Breakdown by departments ----
df_asian_pop <- 
  df %>% 
  select(-pct_vm) %>% 
  filter(group %in% c("sub_asian_all", "asian_pop")) %>% 
  mutate(group = ifelse(group == "sub_asian_all", "Non-Asian population", "Asian population")) %>% 
  pivot_longer(cols = starts_with("pct_"), names_to = "percent", values_to = "value") %>% 
  mutate(Percent = ifelse(percent == "pct_all", "Asian", "Non-Asian")) %>% 
  filter((group == "Asian population" & Percent == "Asian") |
           (group == "Non-Asian population" & Percent == "Non-Asian"))

top_10 <- 
  df_asian_pop %>% 
  filter(year == 2021, group == "Asian population") %>% 
  arrange(desc(population)) %>% 
  head(10)

plot_top_10 <- 
  df_asian_pop %>% 
  filter(`Department / Agency` %in% top_10$`Department / Agency`)

# Functions ----
plot_department <- function(x){
  ggplotly({
    ggplot(x,
           aes(
             x = reorder(`Department / Agency`, population),
             y = population,
             fill = group,
             colour = group,
             text = paste0(
               "Year: ", year,
               "\nGroup: ", group,
               "\nCount: ", format(population, big.mark = ","),
               "\nPercent: ", value, "%"
             )
           )) +
      geom_bar(stat = "identity",
               position = "stack") +
      theme_minimal() +
      scale_y_continuous(labels = comma) +
      scale_colour_manual(values = c(met.brewer("Hiroshige")[1],
                                     met.brewer("Hiroshige", direction = -1)[1])) +
      scale_fill_manual(values = c(met.brewer("Hiroshige")[1],
                                   met.brewer("Hiroshige", direction = -1)[1])) +
      labs(
        x = "Department/Agency",
        y = "Population",
        fill = "Group",
        colour = "Group"
      ) +
      facet_wrap(~ year, ncol = 1) +
      coord_flip()
  }, tooltip = "text")
}
```
## About

This repository contains code to visualize employee salary data from the Government of Canada's Treasury Board, publicly available [here](https://www.canada.ca/en/treasury-board-secretariat/services/innovation/human-resources-statistics/diversity-inclusion-statistics/distribution-public-service-canada-employees-designated-sub-group-department-members-visible-minorities.html).

## The Data

This data is based on self-identification of equity-deserving group and is only available for the core federal public service. In order to obtain the data, the webpage was scraped with the R package `rvest`. Then, `dplyr` and `tidyr` were used to clean it up and plotting was done with: `ggplot2`, `plotly` `scales`, and `MetBrewer`.

We are including the following groups in `Total Asian Population`:

- Chinese
- Japanese
- Korean
- Filipino
- South Asian/ East Indian
- Southeast Asian

Non-White West Asian was not included because it was grouped along with North African and Arab. If needed, this group can be added into the data.

## The Plots

### Total Asian Population

This plot looks at the total count of self-identifying Asian employees.

```{r plot_total}
ggplotly({
  ggplot(asian_pop,
         aes(
           x = year,
           y = total,
           fill = group,
           colour = group,
           text = paste0(
             "Year: ", year,
             "\nGroup: ", group,
             "\nCount: ", format(total, big.mark = ",")
           )
         )) +
    geom_line(group = 1) +
    theme_minimal() +
    scale_y_continuous(labels = comma) +
    scale_colour_manual(values = c(met.brewer("Hiroshige")[1],
                                   met.brewer("Hiroshige", direction = -1)[1])) +
    labs(
      x = "Year",
      y = "Population",
      fill = "Group",
      colour = "Group"
    )
}, tooltip = "text") %>% 
  layout(hovermode = "x unified")
```

### Year-Over-Year Growth

This plot looks at the year-over-year (YOY) growth of self-identifying Asian employees.

```{r plot_yoy}
## Year-over-year growth ----
ggplotly({
  ggplot(yoy,
         aes(
           x = year,
           y = yoy,
           text = paste0(
             "Year: ", year,
             "\nYear-Over-Year Growth: ", yoy, "%",
             "\nCount: ", format(total, big.mark = ",")
           )
         )) +
    geom_line(group = 1,
              aes(colour = group)) +
    theme_minimal() +
    scale_colour_manual(values = c(met.brewer("Hiroshige")[1],
                                   met.brewer("Hiroshige", direction = -1)[1])) +
    labs(x = "Year",
         y = "Year-Over-Year Growth (%)",
         colour = "Group"
    )
}, tooltip = "text") %>% # plotly wrapper
  layout(hovermode = "x unified")
```

Considering these two plots, we can see that though the YOY is comparable between Asian and non-Asian employees, the pre-existing disproportionality between the groups continue to persist.

## Breakdown by departments

To give context, in the 2016 Census 6,095,235 people in Canada reported having Asian origins, representing 17.7% of Canada's population.[^1] At most, ESDC in 2021 had 12.16% of their employees self-identifying as Asian.

[^1]: Source: Statistics Canada. 2017. Canada [Country] and Canada [Country] (table). Census Profile, 2016 Census.

### Top 10 departments

This is a more focused plot which looks at the top 10 departments with the highest Asian population.

```{r breakdown_top, fig.height = 20}
plot_department(x = plot_top_10)
```

### All departments

This is a really busy plot but it shows the breakdown of Asian and non-Asian employees by count.

```{r breakdown_all, fig.height = 50}
plot_department(x = df_asian_pop)
```

## Future endeavours

Moving forward, these data can be further broken down looking more into the departments/agencies, looking at the classification, salary wages, etc. We can also compare the data to the Workforce Availability for a more accurate comparison.

